---
title: "Explore your data: the tidyverse"
subtitle: "The R Statistical Language · Session 2"
format:
  revealjs:
    footer: A. Stamm <aymeric.stamm@cnrs.fr> · The R Statistical Language · Session 2
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
library(tibble)
library(readr)

penguins <- read_csv("penguins.csv", show_col_types = FALSE)

theme_dark <- function(base_size = 11) {
  ggplot2::theme(
    text = ggplot2::element_text(color = "#F1EEE6"),
    axis.text = ggplot2::element_text(color = "#F1EEE6"),
    axis.line = ggplot2::element_line(color = "grey60"),
    panel.background = ggplot2::element_rect(fill = "transparent", colour = NA),
    panel.grid.major = ggplot2::element_line(color = "grey30"),
    panel.grid.minor = ggplot2::element_line(color = "grey25"),
    plot.background = ggplot2::element_rect(fill = "transparent", colour = NA),
    legend.background = ggplot2::element_rect(fill = "transparent", colour = NA),
    legend.key = ggplot2::element_rect(fill = "transparent", colour = NA)
  )
}
```

# Welcome back

## Where we left off {.inv}

In Session 1 you built the core structures of R:

::: {.spread .spread-md .spread-evenly}
- the [console]{.green}, scripts, and R as a [calculator]{.green};
- [vectors]{.green}: the workhorse, and the four ways to subset them;
- [lists]{.amber}: containers of anything, reached with `[[`/`$`;
- [data frames]{.indigo} & [tibbles]{.indigo}: the tables you analyse;
- [reading]{.indigo} & writing data files.
:::

## Today's map {.inv}

::: {.session-arc}
<div class="step s1"><span class="n">1</span>Foundations</div><div class="connector"></div><div class="step s2 current"><span class="n">2</span>Explore your data</div><div class="connector"></div><div class="step s3"><span class="n">3</span>Repeat &amp; package</div>
:::

Session 2 is the **"explore your data"** move of the course: take rows and columns you *stored*, and turn them into understanding.

::: {.spread .spread-md .spread-evenly}
- [shape it]{.green}: the pipe `|>` and `dplyr`/`tidyr`: `filter`, `select`, `mutate`, `arrange`, `group_by`/`summarise`, `pivot_*`;
- [read it]{.amber}: base-R workhorses, then [ggplot2]{.amber}'s layered grammar: histograms, scatters, facets, boxplots;
- [understand it]{.indigo}: the **transform → plot → interpret** loop that is Exploratory Data Analysis.
:::

::: {.callout-desc}
**The thread.** We keep the penguins data the whole session: shape it, then plot it, then read the answers back off the plots. The EDA pipeline you build here is what the *repeat & package* session and the Reproducible Workflows course will build on.
:::

# Tidyverse data transformation

## The tidyverse at a glance {.inv}

A **family of R packages** that work as one: consistent syntax, a shared pipe, and the same *data frame in → data frame out* flow.

::: {.spread .spread-md .spread-evenly}
- [readr]{.green}: bring data *in* (`read_csv`);
- [tibble]{.green}: the modern data frame;
- [dplyr]{.amber}: *transform* rows, columns, groups (**today**);
- [tidyr]{.amber}: *reshape*: keep data [tidy]{.amber} (`pivot_*`);
- [ggplot2]{.indigo}: *visualise* (the very next section);
- plus `purrr` (iterate), `stringr` (strings), `forcats` (factors), `lubridate` (dates).
:::

::: {.callout-desc}
**Today we master `dplyr` + `tidyr`**: the transform engine that feeds every `ggplot` and every analysis. The long tail (`purrr`, `stringr`, …) you pick up on demand; they all share the same one grammar.
:::

## The pipe: `|>` {.inv}

`|>` sends the value on its left into the *first argument* of the function on its right: read it as "**then**":

::: {.split-2}
::: {.codewindow .r width="100%"}
The pipe: top-to-bottom
```r
penguins |> filter(species == "Adelie")
#  reads: penguins, THEN filter(...)
```
:::

::: {.codewindow .r width="100%"}
The nested equivalent: inside-out
```r
filter(penguins, species == "Adelie")
#  order of operations is reversed
```
:::
:::

::: {.paper-card}
The pipe keeps your script readable **top-to-bottom**; the nested alternative reads inside-out. The backbone of data work: one rule, **the data always goes first**.
:::

## Rows and columns: `filter` & `select` {.inv}

Two verbs get you the *slice* you need: **rows** by a condition, **columns** by name:

::: {.codewindow .r width="100%"}
filter rows, select columns
```r
penguins |> filter(body_mass_g >= 4000)              # keep rows that match
penguins |> filter(species != "Chinstrap", !is.na(sex))  # conditions AND together
penguins |> select(species, island, body_mass_g)     # keep these columns
penguins |> select(-bill_depth_mm)                   # …or drop just one
```
:::

::: {.paper-card}
`filter()` keeps **rows**; `select()` keeps **columns**. Multiple conditions in `filter` are AND-ed by default.
:::

## New columns: `mutate` {.inv}

`mutate()` computes **new columns** from existing ones, keeping everything else:

::: {.codewindow .r width="100%"}
Compute a new column with `mutate()`
```r
penguins |>
  mutate(bmi = body_mass_g / (flipper_length_mm / 100)^2) |>
  select(species, flipper_length_mm, body_mass_g, bmi)
```
:::

::: {.paper-card}
New columns are available to later steps in the **same** pipe, and you can `mutate()` several columns at once.
:::

## Recode with `case_when` {.inv}

`case_when()` turns a sequence of **if…else** conditions into a new column: a tidyverse-friendly replacement for nested `ifelse`:

::: {.codewindow .r width="100%"}
Multiple conditions, first match wins
```r
penguins |>
  mutate(size = case_when(
    body_mass_g >= 5000 ~ "large",
    body_mass_g >= 3500 ~ "medium",
    .default = "small"
  ))
```
:::

::: {.paper-card}
`case_when` reads like a **look-up table**: each `condition ~ value` row is evaluated in order, and `.default` catches everything else. Perfect for binning or relabelling inside a `mutate()`.
:::

## Strings with `stringr` {.inv}

`stringr` gives you one consistent family for **text**: every function is `str_*` with the string first, so they all pipe:

::: {.codewindow .r width="100%"}
The `str_*` verbs: detect, replace, extract
```r
penguins |>
  filter(str_detect(species, "tin"))          # species contains "tin" (Chinstrap…)

species_clean <- penguins |>
  mutate(species = str_to_title(species))     # "adelie" → "Adelie"
```
:::

::: {.paper-card}
`str_detect` pairs with `filter`, `str_replace`/`str_to_*` with `mutate`: strings behave exactly like any tidyverse column once you use `str_*`.
:::

## `across()`: apply to many columns {.inv}

`across()` applies one expression to **every column that matches** a selector: no copy-pasting verbs per column:

::: {.codewindow .r width="100%"}
Apply `mean` within a group, over several columns
```r
penguins |>
  group_by(species) |>
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))
```
:::

::: {.paper-card}
`across(where(is.numeric), …)` targets a whole **family** of columns at once: the bridge from "many similar lines" to programmatic thinking you'll meet again in Session 3's functions.
:::

## Sort: `arrange` {.inv}

`arrange()` sorts the rows; wrap a column in `desc()` to go **descending**:

::: {.codewindow .r width="100%"}
Sort with `arrange()` / `desc()`
```r
penguins |> arrange(body_mass_g)                  # lightest first
penguins |> arrange(desc(body_mass_g), species)   # heaviest first, ties by species
```
:::

## Summarise by group {.inv}

`group_by()` + `summarise()` collapse rows into **one row per group**: `group_by` sets the split, `summarise` reduces each group:

```{r}
#| warning: false
penguins |>
  group_by(species) |>
  summarise(n = n(), avg_mass = mean(body_mass_g, na.rm = TRUE))
```

::: {.callout-desc}
`n()` counts rows per group; `na.rm = TRUE` drops missing values inside `mean()`. The result is a small tidy table, ready to plot or pipe onward.
:::

## Reshape: `pivot_longer` & `pivot_wider` {.inv}

[Tidy data]{.amber}: one row per observation, one column per variable. When it isn't, reshape:

::: {.codewindow .r width="100%"}
Wide ↔ long with `pivot_*`
```r
# wide → long (melt several columns back into one key/value pair)
df |>
  pivot_longer(cols = starts_with("bill_"),
               names_to = "measure", values_to = "mm")

# long → wide (spread key/value columns back out)
df |>
  pivot_wider(names_from = measure, values_from = mm)
```
:::

::: {.paper-card}
Every tidyverse tool expects **tidy data**: reshape first, analyse after.
:::

## Your turn: tidyverse {.paper}

::: {.paper-card}
1. Filter **Gentoo** penguins; `select` three columns → `gentoo`.
2. Add `mass_kg = body_mass_g / 1000`.
3. Group by **island**; `summarise` mean & sd of `bill_length_mm`.
4. Heaviest median body mass per species (`arrange` + `desc`).
5. `pivot_longer` a **wide** table (`names_to`, `values_to`).
:::

::: {.callout-desc}
**Lab 1: tidyverse (≈25 min).** Continue in `lab-1-tidyverse.R`. This is the *foundation*: the next lab (ggplot2) and the EDA in the Reproducible Workflows course build directly on it. **Labs 1–2 together are the EDA half of the pipeline**: transform in `lab-1`, plot in `lab-2`.
:::

# Plotting with base R

## Why plot? {.inv}

A plot is the fastest way to *feel* your data. Four numbers (Anscombe's quartet) can look identical in a table and totally different in a plot.

::: {.callout-desc}
**The lesson.** Always look at your data before you summarise it. A single well-chosen plot often tells you more than ten summary statistics: this is the heart of *Exploratory Data Analysis* (explored properly in the Reproducible Workflows course).
:::

## Histograms in base R {.inv}

The simplest first look at one variable: **`hist()`**.

::: {.codewindow .r width="100%"}
A histogram of one variable
```r
hist(penguins$body_mass_g,               # the variable
     breaks = 30,                        # how many bins
     col = "steelblue", border = "white",
     main = "Penguin body mass (g)")      # title
```
:::

::: {.callout-desc}
RStudio/Positron draws the plot in the **Plots** pane, and it appears in your report if you render it. `hist()` is one call, but its defaults are utilitarian.
:::

## Other base-R workhorses {.inv}

::: {.codewindow .r width="100%"}
The rest of the base-R drawer
```r
plot(x, y)          # scatter plot
barplot(table(x))   # counts of categories
boxplot(y ~ g)      # one box per group
curve(sin, 0, 2*pi) # plot a function
```
:::

::: {.paper-card}
Base R can already do a lot, but its look is dated and customisation is verbose: *why* we meet **ggplot2** next.
:::

## The limit of base-R plots {.inv}

Base `plot()` is fine for a quick peek. For **publication-grade** graphics (faceting, legends, themes, layered encodings), it gets unwieldy fast.

That's the gap **ggplot2** fills: a single coherent *grammar*, rather than a grab-bag of functions.

::: {.callout-desc}
You'll implement several of these base-R plots **by hand in lab 2**, then rebuild one in ggplot2 to feel the difference.
:::

# ggplot2

## The grammar of graphics {.inv}

**ggplot2** (part of the tidyverse) builds every plot from a few composable pieces:

::: {.spread .spread-md .spread-evenly}
- **[data]{.green}**: a data frame (usually a tibble);
- **[aesthetics]{.green}** (`aes`): *which variables* map to *which visual channel* (x, y, colour, size, fill, …);
- **[geometries]{.amber}** (`geom_*`): *how* the data are drawn (points, lines, bars, histograms, …);
- **[facets]{.indigo}**: small multiples, splitting by a variable;
- **[scales/themes]{.indigo}**: the polish.
:::

## The first ggplot {.inv}

Every plot starts with `ggplot(data, aes(...))`, then adds layers with `+`:

::: {.codewindow .r width="100%"}
The first ggplot
```r
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(bins = 30)
```
:::

::: {.paper-card}
This is the *same histogram* we drew in base R, but now each piece is explicit and easy to extend.
:::

## Scatter: two continuous variables {.inv}

Mapping variables to `x` and `y`, and a third to `color`:

```{r}
#| fig-width: 7
#| fig-height: 4.5
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species), alpha = 0.7) +
  theme_dark()
```

::: {.callout-desc}
**Read it aloud.** "Body mass grows with flipper length", and each species clusters in its own region of this space. That's a finding you'd never see from `summary()`.
:::

## One variable per facet {.inv}

**Faceting** splits the plot into small multiples: one per level of a variable:

```{r}
#| fig-width: 7
#| fig-height: 4.5
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(bins = 25, fill = "#2E7D6B", color = "#17251F") +
  facet_wrap(~ species, ncol = 1) +
  theme_dark()
```

::: {.callout-desc}
Different species have *different* distributions of body mass: a comparison that's hard to see in one pooled histogram.
:::

## Boxplots for groups {.inv}

Categorical `x` + continuous `y`:

```{r}
#| fig-width: 7
#| fig-height: 4.5
ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.8) +
  theme_dark() +
  theme(legend.position = "none")
```

::: {.callout-desc}
A boxplot shows the median, the box (IQR) and outliers at a glance: ideal for comparing a numeric variable *across groups*.
:::

## Order factors with `forcats` {.inv}

Plots order themselves however the factor levels say: `forcats` (part of the tidyverse) lets you **reorder** those levels so graphs tell the story you want:

::: {.codewindow .r width="100%"}
`fct_reorder` sorts a factor by another variable
```r
penguins |>
  mutate(species = fct_reorder(species, body_mass_g, median, na.rm = TRUE)) |>
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot()
```
:::

::: {.paper-card}
`fct_reorder` puts the *highest median first* on the axis: order becomes part of your message. `fct_relevel`/`fct_lump` fine-tune and collapse rare levels.
:::

## Layers compose {.inv}

An important ggplot2 habit: plots are built from **many layers**. For example, scatter + a smooth trend:

```{r}
#| fig-width: 7
#| fig-height: 4.5
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species), alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "#E8A33D", linewidth = 1) +
  theme_dark()
```

::: {.callout-desc}
`geom_point()` and `geom_smooth()` *layer* on the same plot, each with its own mapping. This layering is what makes ggplot2 so expressive.
:::

## A grammar, not a menu {.inv}

The power is **consistency**: you learn a few verbs (`geom_point`, `geom_bar`, `geom_histogram`, `facet_wrap`, `aes`) and combine them, instead of memorising a different function per plot type.

There's a gallery of extensions (e.g. `{ggpubr}`, `{patchwork}`, `{plotly}` for interactivity).

::: {.callout-desc}
We'll run a *ggplot2 build-up* live: from a bare histogram to a faceted, titled, themed figure in a handful of added lines. Watch how small edits compound.
:::

## Your turn: plotting {.paper}

::: {.paper-card}
1. Load penguins with `read_csv("penguins.csv")`.
2. Base R: scatter `body_mass_g` vs `flipper_length_mm`, by species.
3. ggplot2: same plot with `color = species`.
4. `hist()` of `body_mass_g` + `facet_wrap(~ species)`.
5. Which approach would you reach for?
:::

::: {.callout-desc}
**Lab 2: plotting (≈25 min).** Continue in `lab-2-plotting.R`. Pair-programme: one types, one narrates. **With Lab 1, this completes the EDA half of the pipeline**: transform then plot, exactly the "explore your data" move from the wrap-up.
:::
# Wrap up

## The EDA loop {.dark auto-animate=true background-color="#17251F"}

::: {.big-title data-id="title" style="left:88px; top:150px; width:720px; font-size:72px;"}
Shape,
read,
understand.
:::

::: {.panel data-id="draft"   style="left:840px; top:150px; width:150px; height:360px;"}
:::
::: {.panel data-id="review"  style="left:1006px; top:150px; width:150px; height:170px;"}
:::
::: {.panel data-id="release" style="left:1006px; top:340px; width:150px; height:170px;"}
:::

## Shape, read, understand {.light auto-animate=true background-color="#F1EEE6"}

::: {.big-title data-id="title" style="left:92px; top:104px; width:1100px; font-size:44px;"}
Exploring data is one loop.
:::

::: {.panel .card data-id="draft" style="left:96px; top:220px; width:340px; height:360px;"}
[Shape it]{.h}

[The pipe and `dplyr`/`tidyr`: `filter`, `select`, `mutate`, `arrange`, `group_by`/`summarise`, `pivot_*`, to get the data the shape you need.]{.p}
:::
::: {.panel .card data-id="review" style="left:470px; top:220px; width:340px; height:360px;"}
[Read it]{.h}

[Base-R workhorses (`hist`, `plot`, `boxplot`), then ggplot2's layered grammar: mapping, geoms, facets, themes.]{.p}
:::
::: {.panel .card data-id="release" style="left:844px; top:220px; width:340px; height:360px;"}
[Understand it]{.h}

[Always look before you summarise: a well-chosen plot answers questions ten summary statistics can't.]{.p}
:::

## Understand: the thread {.light auto-animate=true background-color="#F1EEE6"}

::: {.big-title data-id="title" style="left:92px; top:104px; width:1100px; font-size:44px;"}
Transform → plot → interpret.
:::

::: {.panel .card data-id="release" style="left:96px; top:200px; width:800px; height:400px;"}
[The EDA loop]{.h}

[Shape the data until it's tidy, plot it to *see* it, and read the answer back off the plot; then reshape and replot to chase the next question. This transform → plot → interpret loop is the heart of data science: repeat it often, and the *repeat & package* session shows you the loops underneath and turns the moves you repeat into reusable functions you keep.]{.p style="font-size:24px; line-height:1.6;"}
:::

::: {.panel .marker data-id="draft"  style="left:944px; top:230px;"}
:::
::: {.panel .marker data-id="review" style="left:944px; top:290px;"}
:::

## Next time {.dark .finale auto-animate=true background-color="#17251F"}

::: {.big-title data-id="title" style="left:240px; top:300px; width:800px; font-size:66px; text-align:center;"}
Repeat & package: loops, matrices, functions.
:::

::: {.panel data-id="draft"   style="left:470px; top:150px; width:110px; height:110px;"}
:::
::: {.panel data-id="review"  style="left:590px; top:150px; width:110px; height:110px;"}
:::
::: {.panel data-id="release" style="left:710px; top:150px; width:110px; height:110px;"}
:::

::: {style="position:absolute; left:96px; top:600px; width:1024px; font-size:17px; color:rgba(241,238,230,0.62);"}
Between sessions: finish any outstanding labs, and skim [R for Data Science, ch. 3 (visualisation)](https://r4ds.hadley.nz/data-visualize) to reinforce the ggplot2 grammar you'll reuse everywhere.
:::
