---
title: "What does reproducible mean?"
subtitle: "Reproducible Workflows · Session 1"
format:
  revealjs:
    footer: A. Stamm <aymeric.stamm@cnrs.fr> · Reproducible Workflows · Session 1
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
```

# Before the slides: what does "reproducible" mean to you?

## Class brainstorm {.paper}

::: {.paper-card}
**No slides, no spoilers: just you.** In pairs, spend **4 minutes** writing down everything you associate with the word *reproducible*. Don't filter, don't judge:

- What does a *reproducible analysis* look like?
- When is an analysis *not* reproducible?
- Have you ever tried to **re-run** someone else's (or your own!) code months later? What happened?

We'll collect your answers on the board, group them, and the session builds from there.
:::

<div class="paper-adv">
<button type="button" class="paper-adv-btn">Going further</button>
<div class="paper-adv-backdrop" hidden></div>
<div class="paper-adv-modal" hidden>
<div class="paper-adv-head"><strong>Going further</strong><button type="button" class="paper-adv-close" aria-label="Close">&times;</button></div>
<div class="adv-body">
<p><strong>Pair-share (≈10 min total).</strong> 4 min in pairs, then we gather. There are no wrong answers: the point is to surface your prior experience, whatever your level.</p>
</div>
</div>
</div>

## The reproducibility spectrum {.inv}

Your answers cluster along a spectrum: where do your examples sit?

::: {.incremental .highlight-last}
- **Scratch notes**: ideas in a notebook or a chat window; useful in the moment, impossible to re-run.
- **Scripts**: code that *runs* and *re-runs*… on your machine, today.
- **Reports**: prose *and* code together, with results woven into the narrative.
- **Versioned analyses**: code under git, dependencies pinned, so it re-runs next year on another machine.
- **Published analyses**: a report at a reachable URL that **anyone** can open and trust.
:::

::: {.callout-desc .fragment}
The aim of this course is to move every analysis you produce to the right-hand end: **an analysis that lives in a Quarto document, is versioned in git, has its dependencies pinned, and is published at a URL.**
:::

## What reproducible research is: IMRAD {.inv}

Research, distilled, is a short chain of steps:

::: {.incremental .highlight-last}
1. A **question** worth asking.
2. An **experiment** (or study) that collects **data** to answer it.
3. **Analysis** methods, explained *and* applied to that data.
4. **Results** presented and commented.
5. A **discussion** that puts the results and the original question into perspective.
:::

::: {.callout-desc .fragment}
That structure is often called **IMRAD**: introduction, methods, results and discussion. Reproducibility means shipping the *implementation details* of every one of these steps, so a reader can rebuild them from scratch.
:::

# What is reproducibility?

## "Works here" vs "works anywhere" {.inv}

The test is simple. Take an analysis you did last month and ask three questions:

::: {.incremental .spread}
1. **[Will it re-run?]{.green}**: does the code execute from source, with no hidden clicks or hand-fixes?

2. **[Will it re-run *here*?]{.amber}**: does it reproduce on *your* machine tomorrow, when packages have updated?

3. **[Will it re-run *anywhere*?]{.indigo}**: does it reproduce on a colleague's machine, or in a year?
:::

::: {.callout-desc .fragment}
Everything in this course is in service of making each "will it re-run?" an unequivocal **yes**: for you, and for anyone else.
:::

## What breaks reproducibility {.inv}

The enemies we'll learn to fight, one per later session:

::: {.incremental}
- **[The human memory gap]{.green}**: "I clicked through the setup", "I cleaned it by hand in Excel". Code that lives nowhere but the analyst's head.

- **[The moving environment]{.amber}**: `install.packages()` happily updates *globally*, so your analysis breaks when a dependency version silently changes.

- **[The (dis)appearing data]{.indigo}**: reading a local file path only you have; data that can't be re-downloaded.

- **[The vanished machine]{.indigo}**: your laptop dies, your setup is gone.
:::

::: {.callout-desc .fragment}
The *reproducibility habit* answers all four at once: **document your code, pin your environment, source your data, version everything.**
:::

## What reproducible research looks like {.inv}

Here is the shape of every analysis we build this course:

::: {.incremental .highlight-last}
1. **Data** is **sourced** by code (a documented package, an open-data portal), never by manual download.
2. The **analysis lives in a document** (Quarto) that mixes prose and code, so *seeing* the result shows *how* it was made.
3. **Dependencies are pinned** (`uvr`/`uv`, or `renv`), so the document re-runs tomorrow.
4. **Everything is versioned** with git: every change, every step, reversible.
5. **It is published** to a URL others can reach.
:::

::: {.callout-desc .fragment}
You already learned the R side: vectors, data frames, plotting. This course wraps that skill in a **workflow** so your results can be *trusted, shared and rebuilt by anyone, anywhere.*
:::

## The course at a glance {.inv}

Over three sessions we climb from "it works on my machine" to "it's published and reproducible":

::: {.incremental}
- **[Session 1 · Write]{.green}**: your first Quarto documents: prose + code + results, transparent from source.

- **[Session 2 · Source]{.amber}**: source an open-data set (e.g. data.gouv.fr), *pin* it, **version it with git**, and start your **portfolio book**.

- **[Session 3 · Publish]{.indigo}**: deepen git (pull requests, review), manage deps with **uvr/uv**, publish to **Posit Connect**, and finish Chapter 1.
:::

::: {.callout-desc .fragment}
**By the end you'll have a published, versioned, reproducible portfolio**: and the habit of making every future analysis reproducible by default.
:::

# Setting up your environment

## Your toolbox {.inv}

One overarching language (R) plus a small publishing/versioning stack:

::: {.spread .spread-md .spread-evenly}
**[R]{.green}**: the language. Installed from CRAN.

**[Positron]{.green}**: the IDE for both R *and* Python in this course.

**[Quarto]{.amber}**: renders R *or* Python into documents, reports, slides and books: the heart of everything we do.

**[git + GitHub]{.indigo}**: version control; we add this in Session 2.

**[uvr/uv]{.indigo}**: pins the Python (and R) dependencies behind a project, so it re-runs next year on another machine.

**[Posit Connect]{.indigo}**: where your published report goes live at a URL (Session 3).
:::

## Create a Quarto document {.inv}

The simplest way in Positron: **New → New File… → Quarto document**.

::: .spread-cap
A blank `.qmd` opens. Press **Preview** (or Cmd/Ctrl+Shift+K) to render; edit a heading (a line starting with `##`) and the output updates. Most-used Markdown:
:::

- `#/##/###` headings, `**bold**`, `*italic*`, `- ` bullets, `[t](u)` links

::: {.callout-desc}
**Live block (≈15 min).** Together: create a document, render it, and open source (`.qmd`) and preview side by side. See [Markdown basics](https://quarto.org/docs/authoring/markdown-basics.html) for headings, emphasis and lists.
:::

::: {.callout-desc .fragment}
**Your turn (lab 1, steps 0–1, ≈10 min).** Create and render your own document, then write one short prose paragraph about yourself (a heading, a bold or italic word, a bullet list). Download `lab-1-first-doc.qmd` from the course home page; steps 2–5 come later.
:::

## Verify your install {.inv}

Run these and compare outputs:

::: {.codewindow .r name="r-version.R"}
```r
R.version.string   # e.g. "R version 4.6.1"
```
:::

::: {.codewindow .console name="tool-versions.sh"}
```bash
quarto --version
python3 --version
git --version
```
:::

::: {.callout-desc}
**Rule of thumb.** If any of these errors, resolve it *now*: a broken toolchain will surface in every later session. The step-by-step guides are in <https://astamm.github.io/r-ensai/setup.html>.
:::

## Think in terms of projects {.inv}

Open your course folder as a **project** and work there: the working directory *is* that folder.

::: {.split-2}
Every file path in R is resolved **relative to your working directory** unless you write an absolute one. A small, predictable layout keeps a project tidy:

::: {.term}
<span style="color:rgba(241,238,230,0.45);"># a tidy project layout</span>
```text
my-project/
├── data/
│   ├── raw/
│   └── processed/
└── scripts/
```
:::
:::

::: {.incremental .highlight-last}
- In Positron, open the folder as a project (File → Open Folder); a relative path like `read_csv("data/raw/penguins.csv")` then just works.
- **Prefer relative paths**: they let a script run on any machine with the same folder layout.
- Projects are the unit you'll version, pin and publish in Sessions 2–3.
:::

::: {.callout-desc .fragment}
Thinking in projects, not loose files, is the first reproducibility habit: it makes your work portable, and it's the container everything else (git, uvr, publishing) builds on.
:::

# Quarto documents

## What is a Quarto document? {.inv}

A **`.qmd`** file is plain text that mixes three things:

::: {.spread}
[**Markdown for prose**]{.green}: headings, lists, emphasis, links, just like a README.

[**Code *cells***]{.amber} that run R or Python when the document renders, with their output (numbers, figures, tables) written **into** the document.

[**YAML front matter**]{.indigo} at the top (`title`, `format`, …) controlling how the whole thing is rendered.
:::

::: {.callout-desc}
Render it once (`quarto render`, or *Preview* in Positron) and you get a **self-contained report**: the story *and* the evidence, generated from the same source.
:::

## Your first Quarto document {.inv}

Minimal but complete, this *is* a reproducible analysis:

::: {.split-2}
::: {.codewindow .qmd name="report.qmd"}
````markdown
---
title: "My first report"
format: html
---

Here is the mean of `x`:

```{{r}}
x <- c(2, 4, 6, 8, 10)
mean(x)
```

Today is `r "\u0060r Sys.Date()\u0060"`.
````
:::

::: {.qmd-render}
<h1 class="title">My first report</h1>

<p>Here is the mean of <code>x</code>:</p>
<div class="cell">
<div class="sourceCode cell-code htmlmock"><pre class="sourceCode r"><code class="sourceCode r">x <span class="ot">&lt;-</span> <span class="fu">c</span>(<span class="dv">2</span>, <span class="dv">4</span>, <span class="dv">6</span>, <span class="dv">8</span>, <span class="dv">10</span>)
<span class="fu">mean</span>(x)</code></pre></div>
<div class="cell-output cell-output-stdout">
<pre><code>[1] 6</code></pre>
</div>
</div>
<p>Today is 2026-09-09.</p>
:::
:::

::: {.callout-desc}
**Inline code** (`r "\u0060r Sys.Date()\u0060" `) runs R *inside a sentence*; a **code cell** (fenced `{r}`) prints its result below it. Both re-evaluate on every render.
:::

## The YAML header {.inv}

YAML is the control panel at the top of every document:

```yaml
---
title: "My first report"
author: "Your Name"
format: html
---
```

::: {.incremental .highlight-last}
- `title`, `author`: metadata shown at the top of the page.
- `format`: what to build, `html`, `pdf`, `docx`, `revealjs` (these slides are a `revealjs`), or a whole **book** (Session 2).
- Many more **options**: themes, table of contents, captions…
:::

::: {.callout-desc .fragment}
YAML uses strict *key: value* pairs with two-space indentation. Indentation mistakes are the single most common Quarto error: the error message points at the line. Full options: [Front matter](https://quarto.org/docs/authoring/front-matter.html) · formats: [All formats](https://quarto.org/docs/output-formats/all-formats.html).
:::

## Code cells: controlling what shows {.inv}

An **option** is a `#|` comment on its own line inside the cell, e.g. `#| echo: false`:

| Option | Effect |
|---|---|
| `label: fig-…` / `label: tbl-…` | cross-reference a figure or table |
| `eval: false` | keep the code, don't run it |
| `echo: false` | hide the *code*, show the output |
| `include: false` | show neither (setup code) |
| `fig-cap: "…"` / `tbl-cap: "…"` | caption a figure or table |
| `message/warning: false` | silence package-load chatter |

::: {.callout-desc}
By default a cell shows **code + output**; each `#|` option changes that. The full list: [Code execution options](https://quarto.org/docs/computations/execution-options.html).
:::

## One tool, many languages {.inv}

Quarto is **language-agnostic**: the same `.qmd` renders R, Python, or both.

::: {.incremental .highlight-last}
- An R cell and a Python cell sit side by side in one document, each with the same options.
- Which language to use is a property of each **cell**, not of the document.
- R and Python dominate in data science, but Quarto also speaks [**Julia and Observable JS**](https://quarto.org/docs/faq/#what-programming-languages-are-supported-in-quarto), plus more via Jupyter.
- Your portfolio (Session 2) is **Python-compatible by design**: prose plus whichever language fits the analysis.
:::

::: {.callout-desc .fragment}
Render once and you get a single *self-contained report*, whatever the language. That portability is why this course's toolchain covers both R and Python.
:::

## A Quarto document in action {.inv}

Everything so far is easier to *see* than to describe. Try it for yourself:

::: {.spread}
- Open the document you just created in Positron.
- Press **Preview** (or Cmd/Ctrl+Shift+K) to render it.
- Change a heading or a number, save, re-render, and the output updates to match the source.
:::

::: {.callout-desc}
The loop, **edit source → render → see output**, is the whole of Quarto. Everything else (figures, books, publishing) is that loop scaled up.
:::

# Structured reporting & transparent code

## Literate programming {.inv}

The idea at the heart of Quarto (**prose *and* code woven together**) has a name:

::: {.incremental .highlight-last}
- **Literate programming** (Knuth, 1984): an analysis document is written primarily for *humans to read*, with the *code embedded* so the results it discusses appear in place.
- The reader follows your *thinking*; the machine follows your *instructions*.
- The document is **self-documenting**: the evidence sits beside the claim it supports.
:::

::: {.callout-desc .fragment}
Contrast with the *output-first* habit: a completed PDF with no source is only half an analysis. Here, the **source is the analysis**.
:::

## Structure a report for a reader {.inv}

A reproducible report is a *document*, not a dump of output. Structure it:

::: {.incremental .highlight-last}
1. **Question first**: what are you asking, in one sentence?
2. **Data**: where it came from, how it was sourced (the code shows this).
3. **Cleaning decisions**: what you renamed, recoded, dropped, and *why*.
4. **Figures & tables**: each **captioned** and *referred to in the text*.
5. **Findings**: what you found, honestly, limits included.
:::

::: {.callout-desc .fragment}
**The caption habit.** A figure without a caption is a picture; a captioned figure *referred to in prose* is evidence. You'll practice this in Lab 2 and it becomes the backbone of your portfolio chapter.
:::

## Captions & cross-references {.inv}

Quarto numbers and links your figures and tables automatically. This boxplot is a *live* captioned figure (`#| label: fig-iris` and `#| fig-cap: "..."`):

::: {.split-2}

```{r}
#| label: fig-iris
#| fig-cap: "Sepal length across iris species."
ggplot(iris, aes(Species, Sepal.Length)) + geom_boxplot(fill = "#2E7D6B")
```

::: {}
Now refer to it in prose with `@fig-iris`:

> **@fig-iris** shows that setosa sepals are the shortest.

Re-render this deck (or watch the number) and the cross-reference resolves automatically.
:::

:::

::: {.callout-desc}
`@fig-…` / `@tbl-…` are **cross-references**: the number updates automatically if you move figures around. Every good report reads like a conversation with its own figures.
:::

## Your turn: finish lab 1 {.paper}

::: {.paper-card}
**Take 30 min for [lab 1](index.html#subjects-covered), steps 2–5**: download `lab-1-first-doc.qmd` from the course home page (steps 0–1 are done). Each remaining step maps to a slide you've just seen:

- **Step 2, inline code**: *Your first Quarto document*;
- **Step 3, a code chunk with a figure and its caption**: *Code cells* and *Captions & cross-references*;
- **Step 4, a table with a caption**: *Captions & cross-references*;
- **Step 5, a heading structure**: *Structure a report for a reader*.
:::

## The transparency loop {.inv}

Because source *is* the analysis, any reader can ask **"how did you make that?"** and get a real answer.

::: {.incremental}
- This deck ships with a **code popover**: press `c` on any slide to see the source behind it.
- That's the *habit* we want you to internalize for your own work: the analysis answers its own "how".
- When someone asks "where does this number come from?", the answer is **"render the document"**, not "from my head".
:::

::: {.callout-desc .fragment}
**Lab 2 (≈30 min): build a structured report.** You'll write a short reproducible report on the `palmerpenguins` data (with prose, a captioned figure, and a table), following the structure template `lab-2-structured-report.qmd` downloadable from the course home page.
:::

# Project preview: your portfolio book

## The end goal {.inv}

Everything this course points toward is a single artefact: **your portfolio**.

::: {.spread .spread-md .spread-evenly}
**[A Quarto book]{.green}**: not slides, not a single report, but a *book* that grows all semester.

**[Chapter 1 = your EDA]{.amber}**: a real dataset you care about, fetched by code from an open-data portal of your choice, analysed and cleaned.

**[Versioned, pinned, published]{.indigo}**: under git, dependencies locked, reachable at a URL.
:::

::: {.callout-desc}
The scaffold is on the [course home page](index.html#materials) as a downloadable zip: grab it, render it, make it yours. We set up your actual **Chapter 1** in Session 2.
:::

## The semester plan {.inv}

Here's how your book fills up over the term:

::: {.incremental .highlight-last}
1. **Session 1 (today)**: you learn the document mechanics; the book scaffold renders.
2. **Session 2**: you source a dataset by code (any open-data portal, data.gouv.fr or your own country's), clean it, and drop your EDA into **Chapter 1**.
3. **Session 3**: you put the book under **git**, pin it with **renv/uv**, and **publish** it.
4. **Later courses**: other instructors slot their work into the **placeholder chapters**, so the book keeps growing.
:::

::: {.callout-desc .fragment}
By semester's end you hold a published, versioned, reproducible portfolio, and the skill to do it again for any analysis.
:::

## Three moves, one artefact {.dark auto-animate=true background-color="#17251F"}

::: {.big-title data-id="title" style="left:88px; top:150px; width:720px; font-size:72px;"}
Write,
structure,
grow.
:::

::: {.panel data-id="draft"   style="left:840px; top:150px; width:150px; height:360px;"}
:::
::: {.panel data-id="review"  style="left:1006px; top:150px; width:150px; height:170px;"}
:::
::: {.panel data-id="release" style="left:1006px; top:340px; width:150px; height:170px;"}
:::

## Write, structure, grow {.light auto-animate=true background-color="#F1EEE6"}

::: {.big-title data-id="title" style="left:92px; top:104px; width:1100px; font-size:44px;"}
Today you took the first reproducible steps.
:::

::: {.panel .card data-id="draft" style="left:96px; top:220px; width:340px; height:360px;"}
[Document]{.h}

[Your analysis lives in a `.qmd`: YAML, code cells, inline `r`. Source *is* the report.]{.p}
:::
::: {.panel .card data-id="review" style="left:470px; top:220px; width:340px; height:360px;"}
[Structure & review]{.h}

[Captions, cross-references, honest findings: a structured report (your lab-2) that a stranger can follow.]{.p}
:::
::: {.panel .card data-id="release" style="left:844px; top:220px; width:340px; height:360px;"}
[Grow the book]{.h}

[The scaffold preview: a portfolio book that will keep filling all semester, chapter by chapter.]{.p}
:::

## Your Chapter 1 is next {.light auto-animate=true background-color="#F1EEE6"}

::: {.big-title data-id="title" style="left:92px; top:104px; width:1100px; font-size:44px;"}
From a rendering document to a book with content.
:::

::: {.panel .card data-id="release" style="left:96px; top:200px; width:800px; height:400px;"}
[Grow the book: Chapter 1 is next]{.h}

[Every slide today built toward the same artefact: a book that grows. In Session 2 you source a real dataset by code (any open-data portal), clean it, and drop your EDA into Chapter 1 of the scaffold. That chapter is step two toward the versioned, pinned, published portfolio you'll finish in Session 3.]{.p style="font-size:24px; line-height:1.6;"}
:::

::: {.panel .marker data-id="draft"  style="left:944px; top:230px;"}
:::
::: {.panel .marker data-id="review" style="left:944px; top:290px;"}
:::

## Next time {.dark .finale auto-animate=true background-color="#17251F"}

::: {.big-title data-id="title" style="left:200px; top:300px; width:880px; font-size:52px; text-align:center;"}
Session 2: source your data, clean it, and build the book.
:::

::: {.panel data-id="draft" style="left:470px; top:150px; width:110px; height:110px;"}
:::
::: {.panel data-id="review" style="left:590px; top:150px; width:110px; height:110px;"}
:::
::: {.panel data-id="release" style="left:710px; top:150px; width:110px; height:110px;"}
:::

::: {style="position:absolute; left:96px; top:600px; width:1024px; font-size:17px; color:rgba(241,238,230,0.62);"}
Between sessions: make sure `quarto render` works on a trivial document, and skim [Quarto: Getting Started](https://quarto.org/docs/get-started/) for the parts we didn't cover.
:::
