---
title: "Lab 1: source a dataset from an open-data portal"
subtitle: "Reproducible Workflows · Session 2"
author: "Your Name"
format: html
editor: source
---

> ~35 min. Your task: find a dataset you actually care about on an **open-data
> portal of your choice** ([data.gouv.fr](https://www.data.gouv.fr), or your
> own country's official portal), **source it with code** (no manual download),
> and run a quick exploratory analysis in this Quarto document. This same
> dataset will become **Chapter 1** of your portfolio in Lab 2; choose
> something you'd be proud to show all term.
>
> **Tools in this lab:** an open-data portal of your choice, plus the `{dplyr}`,
> `{ggplot2}`, `{janitor}`, `{naniar}`, `{skimr}` packages (see Step 0).

## Step 0: know the packages

```{r}
#| label: setup
#| message: false
library(dplyr)       # data wrangling
library(ggplot2)     # plots
library(janitor)     # clean_names(), tabyl()
library(naniar)      # vis_miss(), gg_miss_var()
library(skimr)       # skim()
```

The **only portal-specific bit** is how you *fetch* the data. Two options;
choose the one that fits your portal:

- **`{rdatagouv}`** (data.gouv.fr search/API). *An option, not a requirement.*
  If you use data.gouv.fr and want it, install from GitHub:

  ```r
  # in the console, once:
  # install.packages("remotes")
  # remotes::install_github("astamm/rdatagouv")
  ```

- **Any portal by URL**: no extra package needed. Most portals (e.g. your
  country's) expose a downloadable file you can grab with `download.file()`.
  Pick the file's direct URL from the portal's web page.

## Step 1: find a dataset

Browse your chosen portal for a topic you genuinely care about: transport,
sport, environment, culture, health…

**If using `{rdatagouv}`** on data.gouv.fr, you can search from R. Use
`type = "dataset"` to narrow to datasets (not reuses or organizations):

```{r}
#| label: search
#| eval: false
res <- search_datasets("air quality", type = "dataset")
res
```

Read the columns to find the dataset's `slug` and a description that matches
your interest. Pick **one**; you'll live with it for the whole term.

> **Pair up.** Spend 5 minutes describing to a classmate *why* you chose this
> dataset. A clear "because…" is the first sentence of your report.

## Step 2: source it with code

The rule is: **the fetch must be reproducible by anyone re-running the
document.** Record the dataset's page URL in your document; that's the seed
of your provenance statement.

**If using `{rdatagouv}`**, get the dataset's metadata, then find a
**resource** (the actual file) you can download by URL:

```{r}
#| label: fetch
#| eval: false
ds <- get_dataset("my-dataset-slug")     # <- your slug here
ds$resources                             # list downloadable files
url <- ds$resources$url[1]               # the first downloadable file
download.file(url, "data/my-data.csv")
```

**Any other portal**: find the direct download link on the dataset's web
page and fetch it the same way:

```{r}
#| label: download
#| eval: false
url <- "https://www.example-portal.org/.../my-data.csv"   # <- the file's direct URL
download.file(url, "data/my-data.csv")
```

> Record the **dataset name, publisher, and URL** in a short sentence in your
> document. That sentence is your *provenance statement*: the reproducible
> habit in action.

## Step 3: quick EDA

Now the familiar loop from the slides:

```{r}
#| label: read
#| eval: false
dat <- readr::read_csv("data/my-data.csv", show_col_types = FALSE)
```

```{r}
#| label: skim
#| eval: false
skim(dat)                     # one-line glimpse
```

```{r}
#| label: names
#| eval: false
dat |> clean_names() |> names()   # are the names already clean?
```

```{r}
#| label: tabs
#| eval: false
dat |> tabyl(<a-categorical-column>)
```

```{r}
#| label: missing
#| eval: false
vis_miss(dat)                 # where is the missingness?
```

## Step 4: record your cleaning decisions

Under a `## Cleaning decisions` heading, write 2–3 sentences:

- what you renamed / recoded / dropped, and **why**;
- how you'll handle missing values, and **why**;
- one thing that surprised you.

A defensible decision, written down, is what makes the difference between a
script and a reproducible analysis.

## Step 5: one figure with a caption

Add **one** short plot that answers a question you actually want answered, with
a real caption (`#| fig-cap:`). Keep it simple: a boxplot, histogram or line
plot over one variable is plenty. Refer to it in a sentence of prose.

## Done: check yourself

- Is the dataset **sourced by code** (`download.file`, with `{rdatagouv}` only
  as an option), not downloaded by hand?
- Is there a **provenance sentence** (dataset, publisher, URL)?
- Do you have a `## Cleaning decisions` paragraph?
- Is there exactly **one captioned figure**, referred to in prose?

> **Stretch.** Turn your cleaning into a *function* or a `dplyr` pipeline kept
> in a separate `.R` file sourced by this document; so the cleaning is
> reusable and the document stays readable.
