---
title: "Accessing data & building the book"
subtitle: "Reproducible Workflows · Session 2"
format:
  revealjs:
    footer: A. Stamm <aymeric.stamm@cnrs.fr> · Reproducible Workflows · Session 2
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(janitor)
library(naniar)
library(skimr)
penguins <- palmerpenguins::penguins   # penguins sourced from CRAN, not a local CSV
```

# Recap: Session 1 → Session 2

## Where we are {.inv}

Last session you learned the document mechanics; today we fill them with real work.

::: {.incremental .highlight-last}
- Session 1: **write**, Quarto documents, structured reports, transparency.
- Session 2 (today): **source**, get a real dataset, clean it, **version it with git**, land your analysis inside the **portfolio book**, and pick your Chapter-1 dataset.
- Session 3: **publish**, pin the environment (uvr), deepen git (PRs, review), put it at a URL, and finish Chapter 1.
:::

::: {.callout-desc .fragment}
Today's payoff: Chapter 1 (the EDA project) of your portfolio book becomes real. Let's trace the *reproducible* data journey end to end.
:::

# The EDA loop & cleaning decisions

## What EDA is for {.paper}

::: {.paper-card}
**Exploratory data analysis (EDA)** is the back-and-forth loop where you learn what the data can tell you:

1. **Get a quick sense**: shapes, types, missingness, ranges.
2. **Look for problems**: wrong types, missing values, inconsistent names/labels, outliers.
3. **Ask & plot**: small targeted plots to answer questions and surface surprises.
4. **Record decisions**: every cleaning choice documented, because a *defensible* choice is a feature.

EDA is *iterative*: you circle through 1–3 several times before any formal modelling. The output that survives into a report is the cleaned data **plus the story of your decisions**.
:::

## The environment at a glance {.inv}

Your IDE is part of the reproducible loop. In Positron:

::: {.incremental .highlight-last}
- **Data pane / variable explorer**: inspect loaded tables and objects without typing.
- **Inline plots**: plots render in the IDE as you iterate.
- But: the *IDE* is for exploring; the *document* is the record. Anything you discover by clicking must end up as **code in a `.qmd`** to be reproducible.
:::

::: {.callout-desc .fragment}
Rule: **explore by clicking, report by code.** The click is fast; the code is what makes it reproducible.
:::

## One-line summaries: `skimr::skim()` {.inv}

A whole data set summarized in one call, the fastest honest "first look":

::: {.codewindow .r name="skim.R"}

```{r}
#| label: skim
#| message: false
#| echo: true
#| output-location: column
penguins |> 
  skimr::skim() |> 
  summary()
```

:::

::: {.callout-desc}
`skim()` shows type, `n_missing`, and mini-distributions per column in one go (here compacted with `|> summary()`). When someone asks "what's in this data?", `skim()` first.
:::

<div class="paper-adv">
<button type="button" class="paper-adv-btn">See the full `skim()` output</button>
<div class="paper-adv-backdrop" hidden></div>
<div class="paper-adv-modal" hidden>
<div class="paper-adv-head"><strong>Full `skim()` output</strong><button type="button" class="paper-adv-close" aria-label="Close">&times;</button></div>
<div class="adv-body">
<p>The unabridged per-column summary, in its usual wide form:</p>
```{r}
#| label: skim-full
#| message: false
#| echo: false
penguins |> 
  skimr::skim()
```
</div>
</div>
</div>

## Cleaning names & counting: `{janitor}` {.inv}

Real-world data arrives with messy names and labels. `{janitor}` tidies them:

::: {.codewindow}

::: {.editor .r name="janitor-clean.R"}
```r
penguins |>
  janitor::clean_names() |>        # lower-case, underscores, no spaces
  dplyr::slice(1:3) |> 
  gt::gt() |> 
  gt::opt_stylize(style = 4)
```
:::

::: {.editor .r .fragment name="janitor-tabyl.R"}
```r
penguins |>
  janitor::tabyl(species) |>          # a nicer table() for one variable
  gt::gt() |> 
  gt::opt_stylize(style = 4)
```
:::

:::

<div class="paper-adv">
<button type="button" class="paper-adv-btn">See the rendered `{janitor}` tables</button>
<div class="paper-adv-backdrop" hidden></div>
<div class="paper-adv-modal" hidden>
<div class="paper-adv-head"><strong>Rendered `{janitor}` tables</strong><button type="button" class="paper-adv-close" aria-label="Close">&times;</button></div>
<div class="adv-body">
```{r}
#| label: janitor-clean
#| message: false
#| echo: false
penguins |>
  janitor::clean_names() |>        # lower-case, underscores, no spaces
  dplyr::slice(1:3) |> 
  gt::gt() |> 
  gt::opt_stylize(style = 4)
```

```{r}
#| label: janitor-tabyl
#| message: false
#| echo: false
penguins |>
  janitor::tabyl(species) |>          # a nicer table() for one variable
  gt::gt() |> 
  gt::opt_stylize(style = 4)
```
</div>
</div>
</div>

::: {.callout-desc .fragment}
`clean_names()` makes column names *syntactic* (safe to reference); `tabyl()` gives counts with percentages. Both make downstream code far more readable.
:::

## Diagnosing missingness: skim + naniar {.inv}

Before choosing what to do, first see *where* and *how much* data is missing. `{skimr}` (a few slides ago) reports `n_missing` per column; `{naniar}` turns missingness into a figure with legible per-column counts, as in @fig-miss-var:

::: {.codewindow .r name="naniar.R"}
```{r}
#| label: fig-miss-var
#| fig-cap: "How much is missing per column, at a glance."
#| fig-width: 9
#| fig-height: 2.8
#| echo: true
#| output-location: column
#| lightbox: true
naniar::gg_miss_var(penguins)
```
:::

::: {.callout-desc}
`gg_miss_var()` plots the number of missing values *per column* with horizontal, readable labels; `vis_miss()` grids the data set to show the *pattern* case by case. [naniar's site](http://naniar.njtierney.com) lists many more missingness helpers.
:::

## Dealing with missing values {.inv}

Once you've diagnosed it, decide how to handle it. Three honest strategies; each is a *documented decision*, not a default:

::: {.incremental .spread}
1. **Drop** with `tidyr::drop_na()`: simplest, but you may lose whole **columns** or many **rows**; only acceptable when the missing cases are few and not systematically different.
2. **Impute** (fill in a plausible value): keeps all cases, but imputed values are *estimated*, not observed. The [MissingData CRAN Task View](https://cran.r-project.org/web/views/MissingData.html) is the curated map of R's methods.
3. **Keep and use `na.rm = TRUE`**: let functions skip `NA` where they support it; cheapest when missingness is sparse and local.
:::

::: {.callout-desc .fragment}
No free lunch: dropping loses information, imputing adds an assumption, `na.rm = TRUE` only defers the decision. Diagnose first (skim + naniar), then choose and *write down why*. Where you can, make that call together with the **domain experts** who know what each `NA` actually means in this data.
:::

## Tying it to Session 1 {.inv}

This is the same cleaning conversation you started in the base-R course with `penguins_messy.txt`:

::: {.incremental .highlight-last}
- There you fixed wrong **types** (`readr` guessing), separators, and missing values by hand.
- Today the **decisions** are the point: `janitor`, `naniar`, `skimr` don't *make* them for you; they *surface* them faster so you can decide well and record why.
:::

::: {.callout-desc .fragment}
The toolkit changed; the habit didn't: **inspect → decide → document.** Every decision you can justify in a sentence is a reproducible decision.
:::

## Your turn: EDA toolkit {.paper}

::: {.paper-card}
Run this on the `penguins` data (loaded at the top), or a data frame of your own:

1. `skim(penguins)`: which columns have missing values? How many?
2. `penguins |> clean_names() |> names()`: do names need cleaning?
3. `penguins |> tabyl(species)`: what are the group sizes?
4. `penguins |> tabyl(species) |> gt::gt() |> gt::opt_stylize(style = 4)`: present those counts as a styled table (the `{gt}` you met in Session 1).
5. `vis_miss(penguins)`: where is the missingness concentrated?
6. Write one sentence: **"I will handle the missingness by … because …"**, that sentence is your reproducible cleaning decision.
:::

<div class="paper-adv">
<button type="button" class="paper-adv-btn">Going further</button>
<div class="paper-adv-backdrop" hidden></div>
<div class="paper-adv-modal" hidden>
<div class="paper-adv-head"><strong>Going further</strong><button type="button" class="paper-adv-close" aria-label="Close">&times;</button></div>
<div class="adv-body">
<p><strong>~12 min.</strong> This is warm-up for Lab 1; stretch by adding a <code>gg_miss_var()</code> plot.</p>
</div>
</div>
</div>

# Sourcing open data

## Open data, anywhere in the world {.inv}

A reproducible analysis needs data it can **re-fetch by code**, and open-data portals everywhere provide exactly that:

::: {.incremental .highlight-last}
- **France**: [data.gouv.fr](https://www.data.gouv.fr), thousands of public datasets, **license-aware**.
- **Your own country**: nearly every government runs an official open-data portal (data.gov… for the US, Canada, UK, Germany, and many more).
- **The rule is the same wherever you look**: a dataset you can fetch *programmatically* is a dataset your analysis can re-produce.
:::

::: {.callout-desc .fragment}
The reproducible move: don't *download* a dataset by hand; write code that *sources* it, so the analysis is re-runnable and the provenance (publisher + URL) is in the document.
:::

## [{rdatagouv}](https://astamm.github.io/rdatagouv/): the French portal from R (an option) {.inv}

Because we teach from l'ENSAI, **data.gouv.fr** is a natural default. Its thousands of public datasets trend toward the **public, transport, environment and culture** topics the French job market cares about, and they're **license-aware**, so the dataset you showcase is safe to publish. [{rdatagouv}](https://astamm.github.io/rdatagouv/) (maintained by your instructor) wraps it:

```{r}
#| label: rdatagouv
#| eval: false
library(rdatagouv)
# search the FRENCH portal for a topic
search_datasets("population", type = "dataset")
# get metadata + resource URLs for a dataset you found
dataset <- get_dataset("my-dataset-slug")
dataset$resources
```

::: {.callout-desc}
Why a natural pick: pick a **topic you care about** that a future recruiter would also find compelling once your portfolio is online. Whatever you choose, [{rdatagouv}](https://astamm.github.io/rdatagouv/) is an *option, not a requirement*: any portal you can query with a URL (e.g. `download.file()`) works just as well for reproducibility.
:::

## Fetching any portal by URL {.inv}

Whatever portal you choose, the reproducible core is the same: a URL your code downloads:

```{r}
#| label: fetch-url
#| eval: false
url <- "https://…/my-open-dataset.csv"   # any open-data portal, any country
download.file(url, "data/my-data.csv")   # a RE-RUNNABLE line of code
```

::: {.incremental .highlight-last}
- **`download.file()` is portal-agnostic**: it doesn't care whether the URL points at data.gouv.fr or your home country's portal.
- Pair it with a **provenance sentence** (dataset, publisher, URL, license) so a reader can *find* and *trust* the source.
- Some portals ship an R/Python client ([{rdatagouv}](https://astamm.github.io/rdatagouv/) here): a convenience, never a requirement.
:::

::: {.callout-desc .fragment}
**Live block (≈10 min).** Together: `rdatagouv::dg_find_datasets(q = "climat", schema_only = TRUE)` *and* browse one non-French portal (e.g. your home country's). Both approaches collapse to the same thing: a URL to download by code.
:::

## Data that ships as an R package {.inv}

Not every dataset needs a portal: many ship *inside* an R (or Python) package, and that's reproducible too. You've used one already, `palmerpenguins::penguins`.

::: {.incremental .highlight-last}
- A **versioned package** is reproducible data: the package is properly released and versioned on CRAN (or PyPI), and its docs and provenance ship with it.
- Because it's a package, your **environment manager** (`uvr`/`uv`) can lock its exact version, so a re-render gets the same data today and next year.
- Good examples: `{gapminder}`, `{nycflights13}`, `{babynames}`, `{Lahman}` (baseball), `{tidycensus}`, `{rdatagouv}`'s returned data.
:::

::: {.callout-desc .fragment}
A packaged dataset is the ultimate reproducible source: **one line to load, an immutable version, documented**. That's why today's analysis can trust `palmerpenguins` without a portal or a `download.file()`.
:::

## Finding data by code: catalogues & portal APIs {.inv}

Beyond your default portal, two more reproducible ways to *find* a dataset by code:

::: {.incremental}
- **Curated catalogues of packaged data**: the [Rdatasets list](https://vincentarelbundock.github.io/Rdatasets/datasets.html) indexes hundreds of datasets shipping in CRAN packages, each loadable in one line; browse the [CRAN package list](https://cran.r-project.org/web/packages/) and search by name.
- **Portal-API packages**, one wrapper per portal, same convenience as `{rdatagouv}`: `{ausOpenData}`, `{opendatatoronto}`, `{SeattleOpenData}`, `{sfOpenData}`, `{laOpenData}`, `{chiOpenData}`, `{nycOpenData}`, `{nysOpenData}`, `{nolaOpenData}`, `{mtaOpenData}`, `{cincinattiOpenData}`, `{ColOpenData}`, `{ctOpenData}`, `{deOpenData}`, `{istat}`, `{istatR}`, `{COVID19}`.
:::

::: {.callout-desc .fragment}
Every route collapses to the same reproducible core: a **packaged** dataset (versioned with its package) or a **portal** you query by code, both with provenance you can record. The portal is optional; the *code you can re-run* is not.
:::

## Lab 1: your dataset {.paper}

::: {.paper-card}
**Lab 1 (≈35 min):** `lab-1-eda-datagouv.qmd`.

1. **Pick a dataset you actually care about**, from any source we've covered: an open-data portal (via `{rdatagouv}`, a portal-API package, or `download.file()`), or a dataset that ships in a package (e.g. from the [Rdatasets catalogue](https://vincentarelbundock.github.io/Rdatasets/datasets.html)).
2. In a Quarto doc, source it **by code** (no manual download), `{rdatagouv}` for data.gouv, `download.file()` for any portal, the package call for a packaged dataset.
3. Run a quick EDA: `skim()`, `clean_names()`, `tabyl()`, `vis_miss()`.
4. Write 2–3 sentences on your *cleaning decisions*.
:::

<div class="paper-adv">
<button type="button" class="paper-adv-btn">Lab 1: your dataset</button>
<div class="paper-adv-backdrop" hidden></div>
<div class="paper-adv-modal" hidden>
<div class="paper-adv-head"><strong>Lab 1: your dataset</strong><button type="button" class="paper-adv-close" aria-label="Close">&times;</button></div>
<div class="adv-body">
<p>This same dataset becomes <strong>Chapter 1</strong> of your portfolio in Lab 2, so choose something you'd be proud to show for the whole term.</p>
</div>
</div>
</div>

# Versioning your data with [{pins}](https://pins.rstudio.com)

## The (dis)appearing data problem {.inv}

Last session you met "the (dis)appearing data": a fetched file can *change* or *vanish* the moment it's off the portal:

::: {.incremental .highlight-last}
- A portal updates a dataset: your analysis was based on the **old version**.
- A raw URL 404s a year later, and your "reproducible" document can't re-run.
- Even locally, `data/my-data.csv` can be overwritten, and git alone doesn't version the *cached* copy well if you don't commit it.
:::

::: {.callout-desc .fragment}
[{pins}](https://pins.rstudio.com) closes that gap: **version your data** the way git versions your code, so a re-render gets the *same* data a reader expects, not whatever is live today.
:::

## [{pins}](https://pins.rstudio.com): pin data to a versioned board {.inv}

[{pins}](https://pins.rstudio.com) stores a data object on a **board** and records **who wrote what, when**:

::: {.codewindow .r name="pins.R"}
```{r}
#| label: pins-setup
#| eval: false
#| echo: true
# a folder board that KEEPS HISTORY (versioned = TRUE)
bd <- pins::board_folder(path = "pins/", versioned = TRUE)

# fetch open data once, then pin it:
dat <- readr::read_csv("https://…/my-open-dataset.csv")
pins::pin_write(board = bd, x = dat, name = "my-dataset")        # snapshot it, versioned & hashed

# later re-reads get the PINNED copy, reproducibly:
pins::pin_read(board = bd, name = "my-dataset")
```
:::

::: {.callout-desc}
`pin_write()` snapshots the data and records the **version + hash**; `pin_read()` returns that exact snapshot. Re-render → same data, every time.
:::

## Where the versioned copy lives {.inv}

A board is just a folder (or a service), and it's fully inspectable:

::: {.incremental .spread}
- `pin_list(board)`: what's been pinned.
- `pin_versions(board, "my-dataset")`: every snapshot, with hashes and dates.
- `pin_read(board, "my-dataset", version = "…")`: roll back to an older snapshot.
:::

::: {.callout-desc .fragment}
The pinned `pins/` folder is committed alongside your code, so the *exact data* your analysis used is reproducible forever, independent of portal changes or broken URLs.
:::

## Fetch once, pin, analyse {.inv}

A reproducible data workflow in three deliberate steps:

::: {.incremental .spread}
1. **Fetch** the open data by URL: provenance recorded.
2. **Pin** it with `pin_write()`: versioned & hashed on a local board.
3. **Analyse from the pinned copy** (`pin_read()`): stable no matter what happens upstream.
:::

::: {.callout-desc .fragment}
Pinning keeps you honest: the *data* you analysed is on record, not just the *code*. Combined with git (code), uvr/uv (environment) and a URL (delivery), it completes the durable portfolio you build in Session 3.
:::

## Beyond data: version your models too {.inv}

The same "pin what you made" idea extends to the *models* you fit, not just the data:

::: {.incremental}
- **[{vetiver}](https://rstudio.github.io/vetiver-r/)**: pins a **model** to a board, so the exact trained model is deployed or re-loaded, versioned & reproducible like your data.
- It pairs with `pin_write()`/`pin_read()`: you already know the mechanics; [{vetiver}](https://rstudio.github.io/vetiver-r/) just knows how to save/load a model's metadata too.
:::

::: {.callout-tip .fragment title="Same habit, new object"}
data → [{pins}](https://pins.rstudio.com), model → [{vetiver}](https://rstudio.github.io/vetiver-r/articles/vetiver.html), code → [git/GitHub](https://github.com)
:::

::: {.callout-desc .fragment}
Saving a trained model by hand usually loses which packages/versions made it. [{vetiver}](https://rstudio.github.io/vetiver-r/) records those alongside the model, so it re-runs and re-deploys anywhere.
:::

## R and Python, side by side {.inv}

Every reproducibility tool you've met has an equivalent in Python, with the same ideas and the same names:

| Role | R | Python |
|---|---|---|
| **Environment** | [uvr](https://nbafrank.github.io/uvr/) | [uv](https://docs.astral.sh/uv/) |
| **Data & artifacts** | [{pins}](https://pins.rstudio.com) | [{pins}](https://rstudio.github.io/pins-python/) |
| **Model versioning/deploy** | [{vetiver}](https://rstudio.github.io/vetiver-r/) | [{vetiver}](https://rstudio.github.io/vetiver-python/) |

::: {.callout-desc}
Your book is Python-compatible by design: whichever language a chapter uses, the reproducible pipeline (source → pin → version → publish) is the same. Only the **environment** row is a preview: you'll actually pin it (uvr/uv) in Session 3, once the whole data-to-publish loop is in place.
:::

# Versioning your code with git

## Git in one habit {.inv}

You version your *data* with `{pins}`; you version your *code* with git, one round at a time:

::: {.codewindow .terminal}
```bash
git status             # your compass: where am I?
git add <file>        # stage the changes you want to keep
git commit -m "msg"   # snapshot with a note
git push              # send the snapshot to GitHub
```
:::

::: {.incremental .highlight-last}
- **[`usethis::use_git()`](https://usethis.r-lib.org)** turns the folder into a git repo; the **Positron git pane** is the clickable version of the loop.
- **Commit often**: small, descriptive commits beat rare mega-commits; `git status` is your compass.
:::

::: {.callout-desc .fragment}
Version your code the way you just versioned your data. `add → commit → push` is the whole habit; everything else is dress-up.
:::

## A repo + a remote {.inv}

Local history is half the story; a **remote** (GitHub) makes it durable and shareable:

::: {.codewindow .r name="usethis.R"}
```r
usethis::use_git()        # make it a git repo
usethis::use_github()     # create + link the remote, push once
```
:::

::: {.incremental}
- Set up a **GitHub account** first, then register a **PAT** with `gitcreds::gitcreds_set()` (mini-lab below); without it, Positron pushes can't authenticate.
- [{usethis}](https://usethis.r-lib.org) runs it all: `use_git()` to init, `use_github()` to link the remote and push once.
- In the **Positron git pane**, the push button sends your commits to GitHub.
:::

::: {.callout-desc .fragment}
A pushed repo is a backup and a way to share; a local-only repo is one laptop failure from gone. Push early, push often.
:::

## .gitignore: what NOT to commit {.inv}

Not everything belongs in history. Tell git to ignore machine-local noise:

::: {.codewindow .terminal}
```bash
cat .gitignore
```
:::

::: {.incremental .highlight-last}
- R: `.Rhistory`, `.RData`, `renv/library/`
- Render output: `_book/`, `*_files/`
- Editor/OS: `.DS_Store`, `.venv/`, `__pycache__/`
:::

::: {.callout-desc .fragment}
Your `portfolio-book/.gitignore` scaffold already covers these. The rule: **commit source, not output or environment.** Source is what rebuilds everything.
:::

## Branches: keep `main` always-working {.inv}

A branch is a second history you can work on without touching `main`:

::: {.codewindow .terminal}
```bash
git checkout -b eda-chapter-1          # make + switch to a branch
# work on it
git commit -am "add chapter 1"         # stage + snapshot
git push -u origin eda-chapter-1       # share the branch early
# each time you re-work it, re-align with main:
git checkout main && git fetch --all --prune && git pull
git checkout eda-chapter-1 && git merge main
```
:::

::: {.incremental .highlight-last}
- **Mirror how you actually work**: try an idea on a branch, keep it if it works; `main` stays always-working.
- In the Positron git pane, the **branch** menu creates and switches branches.
:::

::: {.callout-desc .fragment}
Re-align with `main` often so conflicts stay small; **never push to `main` directly** (Session 3's pull request decides what merges).
:::

## Connect with your instructor {.paper}

::: {.paper-card}
**Mini-lab (~10 min).** Get your book onto GitHub to share and get feedback during the gap week:

1. If you haven't yet, create and store a **PAT**: `usethis::create_github_token()` then `gitcreds::gitcreds_set()`.
2. [`usethis::use_git()`](https://usethis.r-lib.org) then `use_github()` on your portfolio book.
3. Make a real change, commit, push (`git add` / `git commit` / `git push`).
4. Branch `git checkout -b eda-chapter-1`, push it too.

::: {.callout-desc}
By the end of today, your book is versioned, remote, and ready to review.
:::
:::

<div class="paper-adv">
<button type="button" class="paper-adv-btn">Going further</button>
<div class="paper-adv-backdrop" hidden></div>
<div class="paper-adv-modal" hidden>
<div class="paper-adv-head"><strong>Going further</strong><button type="button" class="paper-adv-close" aria-label="Close">&times;</button></div>
<div class="adv-body">
<p>happygitwithr (<a href="https://happygitwithr.com">happygitwithr.com</a>) is the tutorial base: its chapters on git install, connect GitHub, and the commit-push loop cover every step with screenshots. This is how you'll share work and get feedback via GitHub between sessions.</p>
</div>
</div>
</div>

# Quarto books

## What a Quarto book is {.paper}

::: {.paper-card}
A **book** bundles many `.qmd` chapters into one navigable publication:

- `project: type: book` marks the folder; `_quarto.yml` lists **chapters** in order.
- Each `.qmd` is a chapter; they **cross-reference** and share figures/tables.
- Render once → a site with sidebar navigation, TOC, bibliography.

::: {.callout-desc}
You already know the document mechanics: a book is the same thing scaled to many files, wired together in `_quarto.yml`.
:::
:::

## Anatomy of `_quarto.yml` {.inv}

The book is driven by one small file:

::: {.codewindow .yaml}
```yaml
project:
  type: book
book:
  title: "My Data Science Portfolio"
  author: "Your Name"
  chapters:
    - index.qmd
    - chapters/01-EDA.qmd
    - references.qmd
bibliography: references.bib
format:
  html:
    theme: cosmo
```
:::

::: {.callout-desc}
Add a chapter by appending its path to `chapters:`; list order is book order.
:::

## Make it yours: branding {.inv}

The book's look is yours to define. Two files beside `_quarto.yml` control it:

::: {.split-2}

::: {.codewindow .yaml name="_brand.yml"} 
```yaml
# colours & fonts (the zero-CSS on-ramp)
color:
  palette:
    teal:   "#2E7D6B"   # pick your own accents
    amber:  "#E8A33D"
    indigo: "#4A5899"
    ink:    "#17251F"
    paper:  "#F1EEE6"
  primary: teal
  foreground: ink
  background: paper
```
:::

::: {.codewindow .scss name="styles.scss"}
```scss
/* structural CSS beyond the palette */
/* find selectors via Inspect, then tweak */
.sidebar { width: 300px; }
```
:::

:::

::: {.paper-card}
Edit a `"#hex"`, `quarto preview`, look. Match what *you* like, not the course decks.
:::

## Language-agnostic chapters {.inv}

A Quarto book happily mixes languages, and so can your portfolio:

::: {.incremental .highlight-last .spread}
- Chapter files are just `.qmd`; each chapter can run **R, Python, or both**.
- `format: html` and cross-references work identically either way.
- That's why your portfolio stays useful as you grow into Python in later courses.
:::

::: {.callout-desc .fragment}
**Portfolio-first design:** the scaffold already reserves empty chapters for later instructors (`02-semester-2.qmd`, `03-semester-3.qmd`). Your book is built to grow.
:::

## Lab 2: build your book {.paper}

::: {.paper-card}
**Lab 2 (≈35 min):** `lab-2-portfolio-book.qmd`.

1. **Download the starter book**: grab `portfolio-book.zip` from the course materials page (<https://astamm.github.io/r-ensai/reproducible-workflows/#materials>) and unzip it.
2. **Render** it: confirm the starter book builds.
3. **Drop your Lab-1 EDA in as Chapter 1** (`chapters/01-EDA.qmd`).
4. Re-render the whole book; check the sidebar and cross-references.
:::

<div class="paper-adv">
<button type="button" class="paper-adv-btn">Lab 2: your book</button>
<div class="paper-adv-backdrop" hidden></div>
<div class="paper-adv-modal" hidden>
<div class="paper-adv-head"><strong>Lab 2: your book</strong><button type="button" class="paper-adv-close" aria-label="Close">&times;</button></div>
<div class="adv-body">
<p>By the end of this lab you hold the seed of your portfolio, a <em>book</em>, not just a report. Session 3 puts it under version control and publishes it.</p>
</div>
</div>
</div>

# Finding a great dataset

## Beyond data.gouv.fr: OpenData R packages {.inv}

Pick the portal (or package) that gets you to a dataset you *care* about; the reproducibility rule never changes:

::: {.incremental}
- **France**: [data.gouv.fr](https://www.data.gouv.fr) via `{rdatagouv}`.
- **Your country**: nearly every government runs an official portal (data.gov for the US, Canada, UK, Germany…), fetched by URL.
- **R-friendly datasets**: `{tidycensus}` (US Census), `{opendatatoronto}`, `#TidyTuesday` data on GitHub; Kaggle datasets too.
:::

::: {.callout-desc .fragment}
Whatever the source, the bar is the same: you can fetch it **by code** and record its provenance (dataset, publisher, URL, license). If you can download it with one line of code, it's fair game for Chapter 1.
:::

## Choose your dataset before you leave {.paper}

::: {.paper-card}
**Final task of the session (~10 min).** Leave today having chosen your Chapter-1 dataset so Session 3's work block starts fast:

1. **Pick a dataset you actually care about** on data.gouv.fr, your country's portal, or an R-friendly source above.
2. Record its **source by code** (a URL / a package call) and its provenance (publisher, URL, license).
3. Write **one line: the question you want to answer** with it. That question is your IMRAD seed for Chapter 1.
:::

<div class="paper-adv">
<button type="button" class="paper-adv-btn">Going further</button>
<div class="paper-adv-backdrop" hidden></div>
<div class="paper-adv-modal" hidden>
<div class="paper-adv-head"><strong>Going further</strong><button type="button" class="paper-adv-close" aria-label="Close">&times;</button></div>
<div class="adv-body">
<p>Between now and Session 3 there is a week. If you don't finish, keep refining your dataset and question, and commit it to the <code>eda-chapter-1</code> branch you made earlier today; your instructor can already comment on it there.</p>
</div>
</div>
</div>

# Writing for an audience

## A report is read, not just rendered {.inv}

Your EDA will be read by other humans, instructors, classmates, future employers. Write to *them*:

::: {.incremental .highlight-last}
1. **Structure signals intent**: question, data, cleaning, results, findings.
2. **Captions do the explaining**: a reader should grasp a figure from its caption before reading your prose.
3. **Say what you found *and* what you didn't**: honest limits are credibility.
4. **Keep code in service of the story**: the code shows *how*; the prose says *why it matters*.
:::

::: {.callout-desc .fragment}
The reproducibility narrative: because everything is code and captioned, a stranger re-rendering your `.qmd` both **trusts** and **reproduces** your work. That's the product.
:::

## Preparing work others will extend {.inv}

Your chapters are slots other courses will fill. Make that easy:

::: {.incremental .highlight-last}
- Use the **same structure** every time (the 5-section recipe).
- Give **descriptive labels** to figures/tables (`fig-cap` everywhere).
- Keep **setup in one chunk** so new chapters start from a known environment.
- Leave placeholder chapters **in place**: they're the roadmap.
:::

::: {.callout-desc .fragment}
Think of the book as a shared, growing artefact, not a one-off. Every convention you follow today lowers the friction for the next instructor who appends a chapter.
:::

## Recap: the full loop {.inv}

From data to a readable, reproducible chapter, the whole course in one picture:

::: {.incremental .highlight-last}
1. **Source** open data by code (portal of your choice): provenance recorded.
2. **Explore & clean** ({skimr}/{janitor}/{naniar}): decisions documented.
3. **Report** in a structured chapter: captions, cross-refs, findings.
4. **Assemble** into the book: one navigable portfolio.
5. **Version** your work on git: committed, branched, pushed to GitHub (today).
6. *(Session 3)* **Pin & publish**: uvr/uv ({pins} for data), Posit Connect.
:::

::: {.callout-desc .fragment}
Today you've done steps 1–5: your book is sourced, cleaned, assembled, and under version control. Step 6 (Session 3) pins the environment, reviews through a PR, and turns the local book into a shared, durable URL.
:::

## Three moves, one book {.dark auto-animate=true background-color="#17251F"}

::: {.big-title data-id="title" style="left:88px; top:150px; width:720px; font-size:72px;"}
Source,
clean,
build.
:::

::: {.panel data-id="draft"   style="left:840px; top:150px; width:150px; height:360px;"}
:::
::: {.panel data-id="review"  style="left:1006px; top:150px; width:150px; height:170px;"}
:::
::: {.panel data-id="release" style="left:1006px; top:340px; width:150px; height:170px;"}
:::

## Source, clean, build {.light auto-animate=true background-color="#F1EEE6"}

::: {.big-title data-id="title" style="left:92px; top:104px; width:1100px; font-size:44px;"}
Today your book gained its first real chapter.
:::

::: {.panel .card data-id="draft" style="left:96px; top:220px; width:340px; height:360px;"}
[Source by code]{.h}

[Open data fetched by code (any portal) and **pinned** with `{pins}`: provenance (publisher + URL) recorded.]{.p}
:::
::: {.panel .card data-id="review" style="left:470px; top:220px; width:340px; height:360px;"}
[Clean with evidence]{.h}

[`skimr`/`janitor`/`naniar`: decisions documented as *why*, not just *what*, so a stranger can trust them.]{.p}
:::
::: {.panel .card data-id="release" style="left:844px; top:220px; width:340px; height:360px;"}
[Build Chapter 1]{.h}

[Your EDA assembled into the book: captioned, cross-referenced, written for a reader.]{.p}
:::

## Your Chapter 1, committed {.light auto-animate=true background-color="#F1EEE6"}

::: {.big-title data-id="title" style="left:92px; top:104px; width:1100px; font-size:44px;"}
From a local chapter to a durable, shareable book.
:::

::: {.panel .card data-id="release" style="left:96px; top:200px; width:800px; height:400px;"}
[Build Chapter 1: make it durable next]{.h}

[Everything today lives in `01-EDA.qmd` inside your book, **versioned on git** and pushed to GitHub. In Session 3 that chapter becomes published: **uvr/uv** pins its environment and R version (and today's `{pins}` board versions its data), it's reviewed through a **pull request**, then **Posit Connect** gives it a URL. Same book, suddenly a published, reproducible portfolio.]{.p style="font-size:24px; line-height:1.6;"}
:::

::: {.panel .marker data-id="draft"  style="left:944px; top:230px;"}
:::
::: {.panel .marker data-id="review" style="left:944px; top:290px;"}
:::

## Next time {.dark .finale auto-animate=true background-color="#17251F"}

::: {.big-title data-id="title" style="left:200px; top:300px; width:880px; font-size:56px; text-align:center;"}
Session 3: version it, pin it, publish it.
:::

::: {.panel data-id="draft" style="left:470px; top:150px; width:110px; height:110px;"}
:::
::: {.panel data-id="review" style="left:590px; top:150px; width:110px; height:110px;"}
:::
::: {.panel data-id="release" style="left:710px; top:150px; width:110px; height:110px;"}
:::

::: {style="position:absolute; left:96px; top:600px; width:1024px; font-size:17px; color:rgba(241,238,230,0.62);"}
Between sessions: finish `lab-2-portfolio-book.qmd` so your book builds, commit it to GitHub on your `eda-chapter-1` branch, and refine your **Chapter 1** dataset and question. Your instructor can already comment on it there; bring it ready to publish next session.
:::
