---
title: "Version it, pin it, publish it"
subtitle: "Reproducible Workflows · Session 3"
format:
  revealjs:
    footer: A. Stamm <aymeric.stamm@cnrs.fr> · Reproducible Workflows · Session 3
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
```

# Recap: Sessions 1 & 2 → Session 3

## Where we are {.inv}

You have a working, versioned **portfolio book**; today we make it durable and shareable.

::: {.incremental}
- Session 1: **write**, Quarto documents, transparency.
- Session 2: **source**, real data, cleaned, pinned, and under git in your book's Chapter 1.
- Session 3 (today): **deepen git** (PRs, review), **pin the environment** (uvr), **fix randomness** (`{withr}`), **publish** (Posit Connect), and **finish Chapter 1**.
:::

::: {.callout-desc}
Trade "it works on my machine" for: versioned, pinned (version + packages), published at a URL anyone can open.
:::

# Deepen your git

## Why version control (recap) {.inv}

Your smallest reproducible unit is not a file: it's the *history of changes* to a project.

::: {.incremental}
- **Undo anything**: every state of your project is recoverable.
- **Answer "what changed and when?"**: a full log of your analysis's evolution.
- **Collaborate safely**: branches, PRs, review, no clobbering.
- **A backup that lives off-machine**: your laptop dying stops being a catastrophe.
:::

::: {.callout-desc}
Git is the time-machine around your analysis; today we make the habit rigorous enough to share and review.
:::

## Commit, branch, push: already yours {.inv}

You built these reflexes in Session 2; today we sharpen them:

::: {.term}
<div class="prompt">$</div> git status                    # your compass: where am I?
<div class="prompt">$</div> git add <file>                 # stage what you want to keep
<div class="prompt">$</div> git commit -m "message"        # snapshot with a note
<div class="prompt">$</div> git checkout -b <branch-name>  # never work on main
:::

::: {.incremental}
- **Commit often**: small, descriptive commits beat rare mega-commits.
- **Never push to `main`**: a PR is how a change reaches the shared history.
:::

::: {.callout-desc}
`usethis::use_git()` and the Positron source-control pane do all of this with clicks.
:::

## Keep a branch aligned with main {.inv}

Before you touch a branch, fold in whatever `main` gained, so you never diverge silently:

::: {.term}
<div class="prompt">$</div> git checkout main
<div class="prompt">$</div> git fetch --all --prune     # see remote changes
<div class="prompt">$</div> git pull                    # bring main up to date
<div class="prompt">$</div> git checkout <branch-name>
<div class="prompt">$</div> git merge main              # fold main into your branch
:::

::: {.callout-desc}
Do this **before** changing a branch, and your branch tracks `main`: conflicts largely disappear.
:::

## A conflict, and what it looks like {.inv}

Two people change the same line, git can't choose: it marks the file for you.

::: {.split-2}
::: {.term}
<span style="color:rgba(241,238,230,0.45);"># my-analysis.R (merge conflict)</span>
```{r, eval=FALSE}
<<<<<<< upstream
penguins_clean <- janitor::clean_names(penguins)
=======
penguins_clean <- readr::read_csv("raw.csv") |>
                   janitor::clean_names()
>>>>>>> incoming
```
:::

::: {.term}
<span style="color:rgba(241,238,230,0.45);"># a genuine .csv conflict also shows this</span>
```{r, eval=FALSE}
<<<<<<< upstream
394,Adelie,2027
=======
394,Adelie,2026
>>>>>>> incoming
```
:::
:::

::: {.incremental}
- Text between `<<<<<<<` and `>>>>>>>` is what each side proposed; delete the markers and keep the right answer.
- In Positron: use the **inline merge editor** (Source Control → merge) to pick upstream / incoming / both.
:::

::: {.callout-desc}
A conflict just means two edits touched the same place: resolve deliberately, commit, move on.
:::

## Pull requests {.inv}

A PR is "please merge my branch into `main`", reviewed before it lands:

::: {.incremental}
- **Positron way**: the **Source Control** pane → create a pull request via the GitHub extension from your pushed branch.
- **GitHub way**: push the branch, then GitHub's web UI offers "Compare & pull request" from the repo page.
- A PR is the *reviewable unit*: it's where your Chapter 1 gets feedback.
:::

::: {.callout-desc}
Merge to `main` **only via a PR**; `main` stays an always-working, always-reproducible line.
:::

## Review, comment, accept {.inv}

A PR is a conversation, not just a merge button:

::: {.incremental}
- **Assign a reviewer** (your instructor, a classmate) to the PR.
- **Comment on specific lines**: both sides of the diff can leave notes.
- **Request changes** or **approve**: the author updates the branch, pushes, and the PR updates.
- Works the same in Positron (GitHub extension) and on the GitHub web site.
:::

::: {.callout-desc}
Review is how your work becomes reproducible to another human; you'll use this on Chapter 1.
:::

## Your instructor, in the loop {.paper}

::: {.paper-card}
Reserve a branch per idea (e.g. `eda-chapter-1`), push it, and open a **draft PR** when you'd like feedback:

- During the gap and today, the PR is how your instructor comments on and reviews your Chapter 1.
- Address comments, push, and the PR updates; no emailing files around.
:::

<div class="paper-adv">
<button type="button" class="paper-adv-btn">Going further</button>
<div class="paper-adv-backdrop" hidden></div>
<div class="paper-adv-modal" hidden>
<div class="paper-adv-head"><strong>Going further</strong><button type="button" class="paper-adv-close" aria-label="Close">&times;</button></div>
<div class="adv-body">
<p>happygitwithr (<a href="https://happygitwithr.com">happygitwithr.com</a>) is the tutorial this course follows. Its pull-requests chapter walks the whole review loop step by step, with screenshots, for both the Positron and terminal routes.</p>
</div>
</div>
</div>

# Pin your environment: `uvr` (and `renv`)

## The moving-environment problem {.inv}

Your analysis depended on package *versions*, which silently change:

::: {.incremental}
- `install.packages()` updates packages **globally**, and today's tidyverse may differ from last month's.
- The **R version itself** also moves: a book may render today and break next year on a newer R.
- "Works on my machine" becomes "worked on *that* machine, that day".
:::

::: {.callout-desc}
The fix: **pin** the versions your project needs, committed with the code. `renv` pins packages; `uvr` also pins the **R version**.
:::

## renv: the state of the art {.inv}

`renv` gives each project its own library and a lockfile of versions:

::: {.term}
<div class="prompt">$</div> renv::init()       # project-private library + renv.lock
<div class="prompt">$</div> renv::snapshot()   # record current versions
<div class="prompt">$</div> renv::restore()    # rebuild the library FROM the lockfile
:::

::: {.incremental}
- **Commit `renv.lock`**: a machine-readable list of every version.
- **`.gitignore` `renv/library/`**: the cache stays out of git.
:::

::: {.callout-desc}
`renv` is the solid, established tool: it pins packages. `uvr` also pins the R version, so that's our focus.
:::

## uvr: pin R itself, not just packages {.inv}

`uvr` wraps `uv` (the Python manager) so R gets the same "one engine, both languages" discipline:

::: {.incremental}
- **`uvr::init()`** sets up a project manifest + lockfile (a `uv` project).
- **`uvr::r_install("4.6.1")`** installs a project-managed R version.
- **`uvr::r_pin("4.6.1")`** records that version in the manifest, so it's **committed and reproducible**.
- **`uvr::add("tidyverse")`** adds to the manifest, regenerates the lockfile, and installs, all in one call.
:::

::: {.callout-desc}
`r_list()` shows versions; `r_use()` picks one for the session, `r_pin()` records it in the project (that's what survives a fresh clone).
:::

## Install uvr {.inv}

The companion R package from GitHub, then the `uvr` binary:

::: {.term}
<div class="prompt">&gt;</div> pak::pak("nbafrank/uvr-r")   # the uvr R package, from GitHub
<div class="prompt">&gt;</div> uvr::install_uvr()           # installs the uvr binary (cross-platform)
:::

::: {.incremental}
- One-time setup: the R package is the interface; `install_uvr()` puts the engine on your machine.
- It relies on `uv`, the same tool that manages Python later in your book.
:::

::: {.callout-desc}
Installing the R package first, then the binary, keeps the whole `uvr` workflow runnable from R functions inside your project.
:::

## What a lockfile actually pins {.inv}

Both lockfiles are plain text, and that's the point. Open one and it reads like a shopping list:

::: {.callout-desc}
`uv.lock` (via uvr) records each R/Python package with **exact version + hashes**; `renv.lock` records each R package + version + source. Neither stores the packages, only the recipe to rebuild them; `uvr` adds the **pinned R version** on top.
:::

::: {.split-2}
::: {.term}
<span style="color:rgba(241,238,230,0.45);"># uvr / uv.lock (excerpt)</span>
[[package]]
name = "dplyr"
version = "1.1.4"
hash = "sha256:2f3a9..."

<span style="color:rgba(241,238,230,0.45);"># and R itself is pinned too</span>
requires-python = ">=4.6.1"
:::

::: {.term}
<span style="color:rgba(241,238,230,0.45);"># renv.lock (excerpt)</span>
"dplyr": {
  "Version": "1.1.4",
  "Source": "CRAN"
}
:::
:::

::: {.callout-desc}
A lockfile isn't magic; it's an explicit, machine-readable record of what you ran with. Diff it over time to see exactly when an environment changed.
:::

## R and Python, side by side {.inv}

Every reproducibility tool you've met has an equivalent in Python, with the same ideas:

| Role | R | Python |
|---|---|---|
| **Environment** | [`uvr`](https://nbafrank.github.io/uvr/) / [`renv`](https://rstudio.github.io/renv/) | [`uv`](https://docs.astral.sh/uv/) |
| **Data & artifacts** | [`{pins}`](https://pins.rstudio.com) | [`pins`](https://rstudio.github.io/pins-python/) |
| **Model versioning/deploy** | [`{vetiver}`](https://rstudio.github.io/vetiver-r/) | [`vetiver`](https://rstudio.github.io/vetiver-python/) |

::: {.callout-desc}
`uvr` literally runs on `uv`, so R and Python share one engine. Your book is Python-compatible by design, and the reproducible pipeline is the same in both languages.
:::

# Reproducible habits: randomness

## Random numbers break reproducibility {.inv}

Any code that uses randomness gives a different answer on every run, unless told otherwise:

::: {.incremental}
- Sampling, simulations, plot jitter, cross-validation splits: all change a figure or number each render.
- A bare `set.seed()` at the top of a document leaks into the whole session and never restores it.
- Result: the same `.qmd` renders "differently", so a reader can't reproduce your exact figure.
:::

::: {.callout-desc}
A document whose figure changes every render isn't reproducible. Fix the seed **where the randomness happens**, not globally.
:::

## Scope a seed with `{withr}` {.inv}

`{withr}` sets an option for one block and restores it afterward, locally and explicitly:

::: {.codewindow .r width="100%" name="reproducible-figure.qmd"}
```{r}
withr::with_seed(2026, {
  # anything random here, for this block only
  sample(penguins$species, 5)
})

# outside the block, the previous seed state is restored
```
:::

::: {.incremental}
- `withr::with_seed(seed, { ... })`: set the seed for exactly that expression.
- `withr::local_seed(seed)` inside a chunk: same idea, restored when the chunk ends.
- Fix the figure *inside the cell that draws it*; every render then matches.
:::

::: {.callout-desc}
Make randomness intentional and local; your Chapter 1's figures become exactly reproducible without leaking a seed into the session.
:::

# Publish it

## What Posit Connect is {.inv}

A published analysis is one anyone can *open at a URL*, with no local environment required:

::: {.incremental}
- **Posit Connect** hosts R and Python content (books, reports, apps, dashboards) behind a stable URL.
- Publishing pushes your **rendered** book to the server; readers just open the link.
- This is the "anywhere" end of the spectrum: reachable, shareable, current.
:::

::: {.callout-desc}
A URL is the end state of the loop: data → analysis → report → versioned → **published**.
:::

## Pushing a Quarto book {.inv}

Rendering is local and reproducible; *publishing* is a one-step deploy:

::: {.term}
<div class="prompt">$</div> quarto render          # build the book (local, reproducible)
<div class="prompt">$</div> quarto publish connect # push to Posit Connect
:::

::: {.incremental}
- First publish asks you to log in / configure the server once.
- Every later publish updates the same URL: your portfolio stays current.
:::

::: {.callout-desc}
The URL shares work *outside* the notebook: an employer, a teammate, or your future self all see the same, current, reproducible book.
:::

## Proof: reproduce it on a blank machine {.inv}

The real test of everything this course taught: hand someone the *recipe* and see if the book reappears:

::: {.term}
<div class="prompt">$</div> git clone <your-url>              # 1. get the source + history
<div class="prompt">$</div> uvr::restore() (or renv::restore) # 2. rebuild env + R version
<div class="prompt">$</div> quarto render                    # 3. rebuild the book
:::

::: {.incremental}
- **Three commands, no "works on my machine"**: a second person (or your future self) rebuilds your exact book.
- If this *doesn't* work on a fresh clone, you've found a reproducibility gap to fix.
:::

::: {.callout-desc}
Publishing gives a URL to *send*; `clone + restore + render` proves the loop closed.
:::

## The full loop, closed {.inv}

Everything this course taught, in one connected picture:

::: {.incremental}
1. **Write** the analysis in Quarto: source *is* the report.
2. **Source** the data by code: `{pins}` versions the fetched copy.
3. **Version** with git: every step reversible, reviewed via PRs.
4. **Pin** your environment with `uvr`/`uv`.
5. **Deploy** your model with `{vetiver}`, pinned like your data.
6. **Publish** to a URL: reachable and shareable.
:::

::: {.callout-desc}
Miss any link and the chain weakens. Keep all six and your analysis is reproducible by anyone, anywhere, at any time.
:::

# Portfolio handoff

## How your book keeps growing {.inv}

Your Chapter 1 was the seed: now it's part of a living artefact:

::: {.incremental}
- Later courses **append chapters** to your book (`02-semester-2.qmd`, `03-semester-3.qmd`): the placeholders you kept are the roadmap.
- Because everything is versioned, pinned, and published, those additions are themselves reproducible.
- By term's end: a multi-chapter, multi-language, published portfolio.
:::

::: {.callout-desc}
The conventions you followed make it easy for other instructors to build *on your work*.
:::

## Your EDA chapter: the rubric {.paper}

::: {.paper-card}
**How Chapter 1 will be assessed**: use this as your checklist for the work block that follows:

| Criterion | Looks like |
|---|---|
| **Question** | a clear one-sentence motivation up front |
| **Reproducible data** | fetched by code and pinned with `{pins}`; provenance recorded |
| **Cleaning decisions** | documented *why*, not just *what* |
| **Evidence** | captioned figures, each referred to in prose |
| **Honesty** | states what was found *and* what wasn't |
| **Structure** | the 5-section recipe (≈ IMRAD) |
:::

<div class="paper-adv">
<button type="button" class="paper-adv-btn">Going further</button>
<div class="paper-adv-backdrop" hidden></div>
<div class="paper-adv-modal" hidden>
<div class="paper-adv-head"><strong>Going further</strong><button type="button" class="paper-adv-close" aria-label="Close">&times;</button></div>
<div class="adv-body">
<p>A chapter that nails these is a chapter that will stand up in a job interview, not just a graded assignment. The rubric maps onto IMRAD: question ≈ introduction, data + cleaning ≈ methods, figures ≈ results, and your closing reflection ≈ discussion.</p>
</div>
</div>
</div>

## Your turn: finish Chapter 1 (most of the session) {.paper}

::: {.paper-card}
**Work block (the rest of the session).** Turn the EDA you built into a *finished* Chapter 1, then publish it.

1. Apply the rubric: complete the question, cleaning decisions, captions, and findings.
2. **Pin** your data (`{pins}`) and environment (`uvr`); scope randomness with `{withr}`.
3. **Commit to a branch**, push, open a PR, address the instructor's review.
4. **Publish** the book to Posit Connect and open the URL.
:::

<div class="paper-adv">
<button type="button" class="paper-adv-btn">Going further</button>
<div class="paper-adv-backdrop" hidden></div>
<div class="paper-adv-modal" hidden>
<div class="paper-adv-head"><strong>Going further</strong><button type="button" class="paper-adv-close" aria-label="Close">&times;</button></div>
<div class="adv-body">
<p>This is the moment everything this course taught comes together: source by code, clean with evidence, pin data and environment, version by git, review by PR, publish to a URL. Ask the instructor and your neighbours for review as you go; that peer review is part of the grade and part of the habit.</p>
</div>
</div>
</div>

## Models and the Python side {.inv}

Your portfolio may grow models and extra languages: same reproducible pipeline, meeting them at the door:

::: {.incremental}
- **[`{vetiver}`](https://rstudio.github.io/vetiver-r/)** pins a trained **model** + the versions that built it, right beside your data on the same board.
- Where a later chapter ships a model, `{vetiver}` gives it a version to deploy, not just `saveRDS()` and hope.
- **Python twins** for everything: [`uvr`](https://nbafrank.github.io/uvr/) / [`renv`](https://rstudio.github.io/renv/)↔[`uv`](https://docs.astral.sh/uv/), [`{pins}`](https://pins.rstudio.com)↔[`pins`](https://rstudio.github.io/pins-python/), [`{vetiver}`](https://rstudio.github.io/vetiver-r/)↔[`vetiver`](https://rstudio.github.io/vetiver-python/).
:::

::: {.callout-desc}
The rubric's criteria scale to any language: source by code, pin data *and* models, version, publish. A `{vetiver}` model re-deploys anywhere.
:::

## Gallery walk / peer review {.paper}

::: {.paper-card}
**Wrap (≈15 min).** Open **two classmates'** published books and give feedback on:
- the **first figure**: can you read its caption and the story it tells?
- the **cleaning section**: could you justify the same decisions in your own words?
- the **findings**: are they specific and honest?
- the **URL**: does it open and work as a finished publication?

Use the rubric: 2 things they did well, 1 thing to improve.
:::

<div class="paper-adv">
<button type="button" class="paper-adv-btn">Going further</button>
<div class="paper-adv-backdrop" hidden></div>
<div class="paper-adv-modal" hidden>
<div class="paper-adv-head"><strong>Going further</strong><button type="button" class="paper-adv-close" aria-label="Close">&times;</button></div>
<div class="adv-body">
<p>Post your book's URL in the shared space; comment on two classmates' via their PRs or a short note. Peer review is the last reproducibility habit: your work is only reproducible if another human can navigate it, and a URL is what makes that test real.</p>
</div>
</div>
</div>

## Three moves, one career {.dark auto-animate=true background-color="#17251F"}

::: {.big-title data-id="title" style="left:88px; top:150px; width:720px; font-size:72px;"}
Write,
version,
publish.
:::

::: {.panel data-id="draft"   style="left:840px; top:150px; width:150px; height:360px;"}
:::
::: {.panel data-id="review"  style="left:1006px; top:150px; width:150px; height:170px;"}
:::
::: {.panel data-id="release" style="left:1006px; top:340px; width:150px; height:170px;"}
:::

## You built this {.light auto-animate=true background-color="#F1EEE6"}

::: {.big-title data-id="title" style="left:92px; top:104px; width:1100px; font-size:44px;"}
By the end of this course you can turn any analysis into a versioned, pinned, published, reproducible portfolio.
:::

::: {.panel .card data-id="draft" style="left:96px; top:220px; width:340px; height:360px;"}
[Write]{.h}

[Your analysis lives in a Quarto doc: source *is* the report.]{.p}
:::

::: {.panel .card data-id="review" style="left:470px; top:220px; width:340px; height:360px;"}
[Source & Version]{.h}

[Data and models **pinned** (`{pins}`/`{vetiver}`); history kept by git (branches, PRs); environment and R version pinned by uvr.]{.p}
:::

::: {.panel .card data-id="release" style="left:844px; top:220px; width:340px; height:360px;"}
[Publish]{.h}

[The whole thing behind a URL anyone can open, reproducible anywhere.]{.p}
:::

## Your portfolio, published {.light auto-animate=true background-color="#F1EEE6"}

::: {.big-title data-id="title" style="left:92px; top:104px; width:1100px; font-size:44px;"}
From one-off report to a career you keep.
:::

::: {.panel .card data-id="release" style="left:96px; top:200px; width:800px; height:400px;"}
[Every chapter is a reproducible chapter]{.h}

[Everything from this course (the Quarto report you `source()` from, the git history, the `uvr`/`uv` lockfile, the published URL) stacks into one portfolio you keep growing. Bring it to your next course, your internship, your first job: an analysis anyone can open, re-run, and trust. That is the reproducible habit, made visible.]{.p style="font-size:24px; line-height:1.6;"}
:::

::: {.panel .marker data-id="draft"  style="left:944px; top:230px;"}
:::
::: {.panel .marker data-id="review" style="left:944px; top:290px;"}
:::

## Thank you: keep the habit {.dark .finale auto-animate=true background-color="#17251F"}

::: {.big-title data-id="title" style="left:140px; top:300px; width:1000px; font-size:54px; text-align:center;"}
Reproducible is a habit, not a tool.
:::

::: {.panel data-id="draft" style="left:470px; top:150px; width:110px; height:110px;"}
:::
::: {.panel data-id="review" style="left:590px; top:150px; width:110px; height:110px;"}
:::
::: {.panel data-id="release" style="left:710px; top:150px; width:110px; height:110px;"}
:::

::: {style="position:absolute; left:96px; top:600px; width:1024px; font-size:17px; color:rgba(241,238,230,0.62);"}
Keep your portfolio growing: every course adds a chapter, and every chapter is reproducible by default. R · Quarto · git · uvr/uv · Posit Connect. Thank you!
:::
